---
title: "Modeling Techniques"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* BRTs
* GLM
* Linear Regression
* Random Forest
* MaxEnt
* Artificial Neural Networks

### Linear and Logistic Regression ###

* model finds best fit linear line between IV and DV
* Linear regression provides continuous output, while logistic regression is best for discrete classification output (eg. yes/no)

```{r, eval=FALSE}
example_logistic <- logistic_reg() |> set_engine("glm")

example_linear <- linear_reg() |>
  set_engine("lm") |>
  set_mode("regression")
```

### GAMs ###

https://noamross.github.io/gams-in-r-course/

GAMs are a middle ground between simple, highly interpretable models (eg. linear models) and black-box machine learning. They can model complex, nonlinear relationships but still retain clear information on the structure of their predictions.

* GAMs can fit data with **smooths/splines**, which are variably shaped functions
  * smooths are made of many smaller **basis functions**
  * the basis functions are summed together with varying weights
  * this means that a single nonlinear relationship has several parameters, creating a more complex model than something linear










