---
title: "Data Transformation"
output: github_document
---

## dplyr

https://r4ds.had.co.nz/transform.html

Example data set: `nycflights13::flights` contains all flights departing NYC in 2013. 

```{r, message=FALSE}
library(nycflights13)
library(tidyverse)
flights
```

the `dplyr` package has 5 essential functions that are useful for most tasks needed in data manipulation. 

### 1. Filter
`filter(dataframe, ...filtering expressions...)`

* Filters observations by their values
* doesn't alter original dataframe -- must save result to new variable

Note: don't use == for floating point calculations, use `near(a, b)` instead


```{r}
#selecting flights leaving in november or december
#note use of %in% 
filter(flights, month %in% c(11, 12))
```

### 2. Arrange
`arrange(dataframe, ... column names ...)`

* Reorders rows 
* when providing more than one column name, each additional column will be used to break ties in values of preceding columns
* **desc(column_name)** reorders by given column in descending order

```{r}
arrange(flights, desc(dep_delay))
```

### 3. Select
`select(dataframe, ...column names...)`

* Allows you to zoom in on a useful subset of variables
* Example operations
  * **c1:c2** selects columns between c1 and c2
  * **-(.)** excludes any columns in the parentheses
  * **starts_with("str")** matches columns starting with given string
  * **ends_width("str")** matches columns ending with given string
  * **contains("str")** matches columns containing given string
* **rename()** is a variant of select() that can rename columsn without dropping variables note explicitly mentioned
* **everything()** can be used with **select()** if there are columns that would be best at the start of the data frame

```{r}
select(flights, year:day)
rename(flights, tail_num=tailnum)
select(flights, time_hour, air_time, everything())
```

### 4. Mutate
`mutate(dataframe, new_column = . , ...)`

* Adds new columns that are functions of existing columns 
* adds columns at end of dataset
* use **transmute()** if only want to keep new variables

```{r}
#want to see new columns
flights_sml <- select(flights, year:day, ends_with("delay"), distance, air_time)
mutate(flights_sml, gain=dep_delay - arr_delay, speed = distance/air_time * 60)
transmute(flights_sml, hours = air_time/60)
```

### 5. Summarize

* collapses a data frame to a single row
* mostly useful when paired with group_by() 
* can use mean(), median(), sd(), etc. 

```{r}
#Note use of na.rm = TRUE --- removes all null values, as otherwise any null would collapse entire output to null
flights %>% 
  group_by(year, month, day) %>%
    summarize(mean=mean(dep_delay, na.rm = TRUE))
```

* during aggregation, it's useful to include a count, or count of non-missing values, in order to prevent drawing conclusions from small data sets. 


